\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-28 tex/paper.tex
Some cleanup

\begin{abstract}
Complex computer simulators are increasingly used across fields of science as
generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely admit a tractable density or likelihood
function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
inference algorithm for fitting a non-differentiable generative model incorporating ideas from empirical Bayes and variational inference.
We adapt the training procedure of generative
adversarial networks by replacing the differentiable generative network with a
domain-specific simulator. We solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the two adversarial objectives.
Effectively, the procedure results in learning a
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
We present results of the method with simulators producing both discrete and continuous data.

@@

%Computer simulators are used to describe complex data generation processes in a diverse set of fields of science including particle physics, climatology, epidemiology,  and population
%genetics,

In many fields of science such as particle physics, epidemiology,  and population
genetics, computer simulators are used to describe complex data generation processes. These simulators relate
observations $\bfx$ to the parameters $\bftheta$ of an underlying theory or mechanistic model.
In most cases, these simulators are specified as procedural implementations of forward, stochastic processes involving latent variables $\bfz$.
% and may generate either discrete or continuous data.
Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx |
\bftheta)$. The prevalence and significance of this problem has motivated an
active research effort in so-called \textit{likelihood-free inference}
algorithms such as Approximate Bayesian Computation
(ABC) and density estimation-by-comparison algorithms~\cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential,
sisson2011likelihood, marin2012approximate, cranmer2015approximating}.
% Often
% the simulation code involves control-flow that implies the dependence of $\bfx$
% on $\bfz$ is non-differentiable, making inference even more difficult and
% precludes approaches such as Ref.~\citep{2016arXiv160507826G}.

%\kcnote{HMC https://arxiv.org/pdf/1503.01916.pdf}

%\kcnote{add ABC refs mentioned in tex}
%(Beaumont et al., 2002; Marjoram et al., 2003; Sisson et al., 2007; Sisson \& Fan, 2010; Marin et al., 2012; Fan et al., 2013) }

%Often the simulation code involves control-flow that leads to models such that even $p(\bfx | \bftheta)$ is non-differentiable with respect to $\bftheta$, making inference even more difficult and precluding approaches such as Ref.~\citep{2016arXiv160507826G}.

@@

%%In contrast, scientific simulators involve control-flow that leads to models such that $p(\bfx, \bfz | \bftheta)$ is non-differentiable with respect to $\bftheta$, making inference even more difficult and precluding approaches such as Ref.~\citep{2016arXiv160507826G}.
%Scientific simulators can be thought of as highly regularized generative models as they typically have relatively few parameters  endowed with some level of interpretation. In contrast, generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation. In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latent variables $\bfz$.

%While generative models based on neural networks are highly parametrized and the model parameters have no obvious interpretation, scientific simulators typically have relatively few parameters that endowed with some level of interpretation. Thus simulators can be thought of as highly regularized, non-differentiable generative models. In this setting, inference on the model parameters $\bftheta$ is often of more interest than the latents $\bfz$.

%Generative models based on neural networks are differentiable with respect to their high dimensional model parameters $\bftheta$, which  with no particular , differentiable with respect to $\bfz$ and $\bftheta$, and the distribution of the latent $\bfz$
%
%In the context of variational inference, one assumes the the likelihood $p(\bfx | \bftheta)$ is available and posterior $p(\bftheta | \bfx)$ is tractable up to a constant (ie. ($p(\bfx, \bfz | \bftheta)$ is tractable). Moreover, one is typically interested in estimating the posterior distribution of the latent variables instead of the parameters of the simulator $\bftheta$.

%In the context of variational inference, these technihave also been used as the recognition model ~\citep{2015arXiv150505770J, DBLP:journals/corr/KingmaSW16, DBLP:conf/icml/RezendeMW14}; however, in those cases the likelihood $p(\bfx | \bftheta)$ is available and posterior $p(\bftheta | \bfx)$ is tractable up to a constant.

%Because it is usually computationally intractable, most simulators
%
%implicit generative models~\citep{2016arXiv161003483M}
%
%In most
%cases, these implicit generative models~\citep{2016arXiv161003483M} are
%specified as procedural implementations of forward stochastic processes that
%generates discrete or continuous data, and rarely do these simulators admit a tractable density. Because it is usually computationally intractable, most simulators
%do not provide a way to directly evaluate the likelihood function for a
%given observation, thereby making inference difficult.
%In addition, most scientific simulators are written using opaque low-level programming
%languages, hence raising the bar for likelihood-free inference
%algorithms relying e.g. on the simulator being a function with computable derivatives~\citep{2016arXiv160507826G}
%or being a controllable probabilistic program~\citep{2016arXiv161009900L}.

In this note, we develop a likelihood-free inference algorithm
for non-differentiable, implicit generative models.
% The first allows us
% to perform empirical Bayes through variational inference, and the
% second provides a point estimator of the parameters $\bftheta$.
We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives.
The objective of both algorithms is to match marginal distribution of
the generated data to the empirical distribution of the observations.

%The procedure results in learning an arbitrarily tight
%proposal distribution over simulator parameters, such that the corresponding
%marginal distribution of the generated data matches the observations.

%In this note, we develop an unsupervised learning algorithm for the point
%estimation of the parameters $\bftheta$ of a non-differentiable, implicit
%generative model. We adapt the adversarial
%training procedure of generative adversarial
%networks~\cite{goodfellow2014generative} by replacing the implicit generative
%network with a domain-based scientific simulator, and solve the resulting
%non-differentiable minimax problem by minimizing variational upper
%bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
%objectives.
%The procedure results in learning an arbitrarily tight
%proposal distribution over simulator parameters, such that the corresponding
%marginal distribution of the generated data matches the observations.

%In this note, we develop a likelihood-free inference algorithm for the point
%estimation of the parameters $\bftheta$ of a non-differentiable, implicit
%generative model. We adapt the adversarial
%training procedure of generative adversarial
%networks~\cite{goodfellow2014generative} by replacing the implicit generative
%network with a domain-based scientific simulator, and solve the resulting
%non-differentiable minimax problem by minimizing variational upper
%bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
%objectives. The procedure results in learning a proposal distribution over
%simulator parameters, hence producing an arbitrarily tight family of models whose
%marginal distribution of generated data matches the observed data.

@@

We consider a family of parametrized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process
%, such that
%\begin{equation}\label{eqn:p_x}
%    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
%\end{equation}
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness.
Unlike implicit generative models defined by neural networks, we do not assume
$\bfz$ to be a fixed-size vector with a simple density. Instead, the
dimension of $\bfz$ and the nature of its components (uniform, normal,
discrete, continuous, etc.) are inherited from the control flow of the
simulation code and may depend on $\bftheta$ in some intricate way. Moreover,
the dimension of $\bfz$ may be much larger than the dimension of
$\bfx$.
%Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.

%Unlike implicit generative models defined by neural networks, we do not assume $\bfz$ to
%be a fixed-size vector (e.g., it can be a sequence of variable length) and its
%distribution $p(\bfz|\bftheta)$ may itself depend on $\bftheta$ in some intricate way
%inherited from the control flow of the simulation code.

We assume that the stochastic generative process that defines $p(\mathbf{x}|\bftheta)$ is
specified through a non-differentiable deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
\mathbb{R}^d$. Operationally, %That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
\end{equation}
such that the density $p(\mathbf{x}|\bftheta)$ can be
written as
%considered as a non-differentiable reparametrization of the random variable $\bfz$
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g_i(\bfz;\bftheta) \leq x_i\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
\begin{equation}\label{eqn:p_x_sim}
    p(\mathbf{x}|\bftheta) = \int_{\{\bfz:g(\bfz;\bftheta) = \bfx \}} p(\bfz|\bftheta) \mu(d\bfz),
\end{equation}
where $\mu$ is a probability measure.
%Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
%used to generate data in forward mode. For this reason, evaluating the integral
%in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
%in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
%e.g. as when $g$ is specified as a computer program.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters
$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the implicit model $p(\mathbf{x}|\bftheta)$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
\end{equation}
where $\rho$ is some distance or divergence.
%~\citep{bernton2017inference}.
%For many choices of divergence $\rho$, this is asymptotically equivalent to a maximum likelihood estimate if $ \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta^*)) \to 0$.

%By plugging Wp in place of D in Eqs. (1) and (2), we obtain the minimum Wasserstein estimator (MWE) and minimum expected Wasserstein estimator (MEWE) of order p, denoted ??n and ??n,m respectively. Some properties of the MWE have been studied in Bassetti et al. (2006), for well-specified models and i.i.d. data; we propose new results in Section 4. Intuitively, under some conditions we can expect ??n to converge to ??, in the sense that Wp(??n,??) ? 0. Consequently, the minimum of ?  ? Wp(??n,??) might converge to the minimum of ?  ? Wp(??,??), denoted by ??, assuming its existence and unicity. In the well-specified case, ?? coincides with the data-generating parameter. In the misspecified case, ?? is typically different from the limit of the maximum likelihood estimator (MLE), which is the minimizer of KL(??|??).

@@

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, with
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\equiv \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
%
%\begin{align}
%    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& - {\cal L}_W
%\end{align}
%\begin{align}
%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]  \nonumber \\
%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)]
%\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick its gradient can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
 %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, robust to noise
and parameter scaling).
% We will use
% Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to make  use instead of the Natural Evolution Strategy
% algorithm~\citep{2011arXiv1106.4487W} or of variance reduction techniques
% such as Rao-Blackwellization,
% Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the
%gradients may be too large to be useful in many problems. Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.

% In this simple form, stochastic gradient descent based on
% Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
% of scale invariance. To avoid the dependence on the parametrization of the
% proposal distribution, natural gradients~\citep{amari1998natural} can be used
% instead to point towards the steepest descent in the space of realizable
% proposal distributions, rather than in the space of their parameters. Effectively,
% the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
% corrects the gradient of $U$ according to the local curvature of the KL-divergence
% surface of proposal distributions, which results in forming
% \begin{equation}\label{eqn:approx-grad-nat}
%     \tilde{\nabla}_\bfpsi U(\bfpsi) = \mathbf{F}(\bfpsi)^{-1} \nabla_\bfpsi U(\bfpsi),
% \end{equation}
% where $\mathbf{F}(\bfpsi)$ is the Fisher information matrix
% \begin{equation}
%     \mathbf{F}(\bfpsi) = \mathbb{E}_{\bftheta \sim q_\bfphi}[ \nabla_\bfpsi \log q_\bfpsi(\bftheta) \nabla_\bfpsi \log q_\bfpsi(\bftheta)^{\top} ]
% \end{equation}
% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\bfpsi \log q_\bfpsi(\bftheta)$ already
% evaluated to form $\nabla_\bfpsi U(\bfpsi)$.

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of $g$ and alternately minimize the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\bfphi$ and $\bfpsi$.
When updating
$\bfphi$, unbiased estimates of $\nabla_\bfphi U_d$ can be obtained by
directly evaluating the gradient of $U_d$ over mini-batches of real and
generated data, as is ordinarily done in stochastic gradient descent. When updating
$\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
That is,
%\begin{equation}\label{eqn:grad-ug-approx}
%   \nabla_\bfpsi U_g = \mathbb{E}_{\subarray{l}{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta) }}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation}
\begin{equation}\label{eqn:grad-ug-approx}
   \nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \tilde{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \tilde{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation}
%\begin{equation}\label{eqn:grad-ug-approx}
%    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation}
which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \bfphi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.
%
%Obviously, the variational relaxation could similarly be coupled with
%other variants of GANs and/or of evolution strategies.

%Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
%have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
%a distribution parametrized in terms of $\bfpsi$:
%\begin{equation}\label{eqn:p_psi}
%    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
%\end{equation}
%Intuitively, this corresponds to a family of simulators, each configured
%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
%More formally, the learned model  therefore corresponds to the marginal distribution
%of the generated data. That is,
%\begin{equation}
%    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
%\end{equation}
%Intuitively, this corresponds to a family of simulators, each configured
%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.

% Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
% variance of the noisy estimator of the gradients may be too large to be useful in many problems.
% Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
% Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
% Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
% empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
% %Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
% These variants may be particularly useful for small $N$.

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems.
We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity use instead the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W} or variance reduction techniques
such as Rao-Blackwellization,
Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.

\subsection{Point Estimates through Marginal Likelihood}

%\subsection{Empirical Bayes through  Variational Inference}

@@

%where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
%%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
%%posterior predictive distribution.

>>>>

\begin{abstract}

Complex computer simulators are increasingly used across fields of science as
generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely admit a tractable density or likelihood
function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
inference algorithm for fitting a non-differentiable generative model incorporating ideas from empirical Bayes and variational inference.
We adapt the training procedure of generative
adversarial networks by replacing the differentiable generative network with a
domain-specific simulator. We solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the two adversarial objectives.
Effectively, the procedure results in learning a
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
We present results of the method with simulators producing both discrete and continuous data.

@@

In many fields of science such as particle physics, epidemiology,  and population
genetics, computer simulators are used to describe complex data generation processes. These simulators relate
observations $\bfx$ to the parameters $\bftheta$ of an underlying theory or mechanistic model.
In most cases, these simulators are specified as procedural implementations of forward, stochastic processes involving latent variables $\bfz$.
Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx |
\bftheta)$. The prevalence and significance of this problem has motivated an
active research effort in so-called \textit{likelihood-free inference}
algorithms such as Approximate Bayesian Computation
(ABC) and density estimation-by-comparison algorithms~\cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential,
sisson2011likelihood, marin2012approximate, cranmer2015approximating}.

@@

In this note, we develop a likelihood-free inference algorithm
for non-differentiable, implicit generative models.
We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives.
The objective of both algorithms is to match marginal distribution of
the generated data to the empirical distribution of the observations.

@@

We consider a family of parametrized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness.
Unlike implicit generative models defined by neural networks, we do not assume
$\bfz$ to be a fixed-size vector with a simple density. Instead, the
dimension of $\bfz$ and the nature of its components (uniform, normal,
discrete, continuous, etc.) are inherited from the control flow of the
simulation code and may depend on $\bftheta$ in some intricate way. Moreover,
the dimension of $\bfz$ may be much larger than the dimension of
$\bfx$.

We assume that the stochastic generative process that defines $p(\mathbf{x}|\bftheta)$ is
specified through a non-differentiable deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
\mathbb{R}^d$. Operationally, %That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
\end{equation}
such that the density $p(\mathbf{x}|\bftheta)$ can be
written as
\begin{equation}\label{eqn:p_x_sim}
    p(\mathbf{x}|\bftheta) = \int_{\{\bfz:g(\bfz;\bftheta) = \bfx \}} p(\bfz|\bftheta) \mu(d\bfz),
\end{equation}
where $\mu$ is a probability measure.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters
$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the implicit model $p(\mathbf{x}|\bftheta)$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
\end{equation}
where $\rho$ is some distance or divergence.

@@

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\equiv \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick its gradient can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
 %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, robust to noise
and parameter scaling).

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of $g$ and alternately minimize the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\bfphi$ and $\bfpsi$.
When updating
$\bfphi$, unbiased estimates of $\nabla_\bfphi U_d$ can be obtained by
directly evaluating the gradient of $U_d$ over mini-batches of real and
generated data, as is ordinarily done in stochastic gradient descent. When updating
$\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
That is,
\begin{equation}\label{eqn:grad-ug-approx}
   \nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \tilde{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \tilde{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation}
which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \bfphi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems.
We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W} or variance reduction techniques
such as Rao-Blackwellization,
Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.

\subsection{Point Estimates through Marginal Likelihood}
%\subsection{Empirical Bayes through  Variational Inference}

@@
